\documentclass[sigconf]{acmart}
\usepackage{graphicx}
\usepackage{algorithmic}
\usepackage{algorithm}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmConference[Algorithms Conference '25]{Proceedings of the Algorithms Conference '25}{May 20-June 3,
  2025}{Ljubljana, Slovenia}
\acmISBN{978-1-4503-XXXX-X/25/05}
\acmDOI{10.1145/XXXXXXX.XXXXXXX}

\acmArticleType{Review}

\title{Verifying Groups in Linear Time}

\author{Vanja Stojanovi\'c}
\email{vs66277@student.uni-lj.si}
\authornote{Corresponding author.}
\affiliation{
  \institution{University of Ljubljana}
  \department{Faculty of Mathematics and Physics}
  \city{Ljubljana}
  \country{Slovenia}
}

\author{Bor Panger\v si\v c}
\affiliation{
  \institution{University of Ljubljana}
  \department{Faculty of Mathematics and Physics}
  \city{Ljubljana}
  \country{Slovenia}
}

%— Authors from FRI —%
\author{Pia Sotlar}
\affiliation{
  \institution{University of Ljubljana}
  \department{Faculty of Computer Science and Informatics}
  \city{Ljubljana}
  \country{Slovenia}
}

\author{Klemen Kav\v ci\v c}
\affiliation{
  \institution{University of Ljubljana}
  \department{Faculty of Computer Science and Informatics}
  \city{Ljubljana}
  \country{Slovenia}
}

\keywords{Groups, Computational group theory, Cayley tables}

\begin{abstract}
  \bfseries
  We explore the latest upper and optimal bounds on the problem of deciding 
  whether a given \(n \times n\) multiplication table is a Cayley multiplication table of a 
  finite group, presented in \cite{10756141} and its cofindings. Exploring the applications of
  this algorithm, such as in computational group theory software, coding theory \cite{zurekCodingTheory} etc.
  \normalfont
\end{abstract}


\begin{document}

\maketitle


\section{Introduction}
\subsection{Motivation}
Verifying whether a given multiplication table defines a \textbf{group} is a fundamental problem in computational algebra with applications in cryptography, error-correcting codes, and symbolic computation. The key challenge lies in efficiently checking the \textbf{associativity axiom}, which naively requires testing all possible triplets \((a, b, c)\) to ensure:
\[
(a \cdot b) \cdot c = a \cdot (b \cdot c).
\]
A brute-force approach would require \(O(n^3)\) time, which is impractical for large \(n\). While prior work has improved this to \(O(n^2 \log n)\) deterministically or \(O(n^2)\) probabilistically, the question remains: \textbf{Can we do better?}

\subsubsection*{Why Linear Time is Non-Obvious}
A natural hope might be that associativity could be verified in \textbf{subcubic (or even linear) time} by exploiting algebraic structure. However, as demonstrated by the following theorem:

\begin{theorem}[Local Non-Associativity]
On every set with at least 4 elements, there exists a binary operation that is associative everywhere except on one triplet.
\end{theorem}
\begin{proof}
Let \( S \) be a set with \( |S| \geq 4 \). Choose distinct elements \( a, u, v, w \in S \). Define a binary operation \( * \) on \( S \) as follows:
\[
x * y =
\begin{cases}
u & \text{if } x = a \text{ and } y = a, \\
v & \text{if } x = a \text{ and } y = u, \\
w & \text{otherwise}.
\end{cases}
\]
We claim \( * \) is associative except for the triplet \( (a, a, a) \).
\\
\textbf{Non-associativity at \( (a, a, a) \):}
\[
(a * a) * a = u * a = w, \quad \text{but} \quad a * (a * a) = a * u = v.
\]
Since \( w \neq v \), \( (a * a) * a \neq a * (a * a) \).
\\
\textbf{Associativity for all other triplets:}
\\
First, let's analyze LHS $=(x \ast y) \ast z$. Let $p_1 = x \ast y$.
\begin{itemize}
    \item If $(x,y) = (a,a)$, then $p_1 = u$. LHS $= u \ast z$.
    Since $u \neq a$, the pair $(u,z)$ is neither $(a,a)$ nor $(a,u)$. So, $u \ast z = w$ by definition (3).
    \item If $(x,y) = (a,u)$, then $p_1 = v$. LHS $= v \ast z$.
    Since $v \neq a$, the pair $(v,z)$ is neither $(a,a)$ nor $(a,u)$. So, $v \ast z = w$ by definition (3).
    \item If $(x,y) \neq (a,a)$ and $(x,y) \neq (a,u)$, then $p_1 = w$. LHS $= w \ast z$.
    Since $w \neq a$, the pair $(w,z)$ is neither $(a,a)$ nor $(a,u)$. So, $w \ast z = w$ by definition (3).
\end{itemize}
In all possible cases for $(x,y)$, LHS $=(x \ast y) \ast z = w$.
\\
Next, let's analyze RHS $= x \ast (y \ast z)$. Let $p_2 = y \ast z$.
\begin{itemize}
    \item If $(y,z) = (a,a)$, then $p_2 = u$. RHS $= x \ast u$.
    \begin{itemize}
        \item If $x = a$, this corresponds to the triplet $(a,a,a)$. As shown before, RHS $= a \ast u = v$.
        \item If $x \neq a$, the pair $(x,u)$ is not $(a,a)$ (since $x \neq a$) and not $(a,u)$ (since $x \neq a$). So, $x \ast u = w$ by definition (3).
    \end{itemize}
    \item If $(y,z) = (a,u)$, then $p_2 = v$. RHS $= x \ast v$.
    Since $v \neq a$ and $v \neq u$ (as $a,u,v,w$ are distinct), the pair $(x,v)$ cannot be $(a,a)$ (as $v \neq a$) and cannot be $(a,u)$ (as $v \neq u$). So, $x \ast v = w$ by definition (3).
    \item If $(y,z) \neq (a,a)$ and $(y,z) \neq (a,u)$, then $p_2 = w$. RHS $= x \ast w$.
    Since $w \neq a$ and $w \neq u$ (as $a,u,v,w$ are distinct), the pair $(x,w)$ cannot be $(a,a)$ (as $w \neq a$) and cannot be $(a,u)$ (as $w \neq u$). So, $x \ast w = w$ by definition (3).
\end{itemize}

\end{proof}

This construction shows that:
\begin{itemize}
    \item \textbf{A single inconsistency suffices to break associativity.}
    \item \textbf{No local or sparse testing suffices}—even if almost all triplets are associative, the operation may still fail to be a group.
\end{itemize}

\subsection{Problem Statement}
Cayley table describes a structure of a finite group. It arranges all possible products of all the group's elements in a square table. To build a Cayley table we are given a set S and a binary operation \(\cdot : S \times S \rightarrow S\). In case of (\(|S| = n\)), a Cayley table is represented by a \(n\times n\) table, where entry (i,j) is a result of a binary operation on element in row i and element in column j, which is also an element of set S.
To know if a table represents a Cayley table, three properties have to be satisfied:
\begin{enumerate}
    \item Existence of an identity element: \(e: e \cdot s = s \cdot e = s\) for every \(s \in S\).
    \item Existence of inverses: for each \(s \in S\), there exists \(h \in S\) such that \(s \cdot h = h \cdot s = e\).
    \item Associativity: \((a \cdot b) \cdot c = a \cdot (b \cdot c)\) for all \(a,b,c \in S\).
\end{enumerate}
So if we are given an \( n \times n \) multiplication table, the goal is to determine if it represents a valid group.

A simple example of a Cayley table is one for the group \(\{1,-1 \}\):
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
$\times$ & 1 & $-1$ \\
\hline
1   & 1   & $-1$ \\
\hline
$-1$ & $-1$ & 1   \\
\hline
\end{tabular}
\end{center}

Some of the features of a Cayley table are:
\begin{itemize}
    \item Every row and column of the table should contain each element exactly once.
    \item The identity element should appear in every row and column exactly once and it should be "distributed" on the main diagonal.
    \item Each entry must also be a row/column label. Otherwise, the operation is not closed.
    \item In the table there should be one row in which the column labels appear in order, this indicates the presence of an identity element.
\end{itemize}

\subsection{Applications of the Finding and Usage of Cayley Tables}

Within computational group theory software, such as GAP and Magma, efficient Cayley table verification plays a foundational role.
These systems allow users to define finite groups by providing their Cayley tables.
Upon input, the software must verify if the given table adheres to the group axioms (closure, associativity, identity, and inverses)
before allowing further computations or analysis. 

For instance, if a user intends to find the subgroups, conjugacy classes, or
other properties of a group defined by its Cayley table, the system first needs to ensure that the input indeed represents a valid group. 
Efficient verification algorithms make this initial validation process practical, especially for groups of moderate size. 
Furthermore, while not the most scalable approach for very large groups, Cayley tables can be employed in group isomorphism 
testing, particularly for groups with a smaller number of elements \cite{williams2015group}.

Finite groups also play a role in coding theory, particularly in the construction of error-correcting codes. For example, cyclic codes have a close 
relationship with cyclic groups. The algebraic properties of the underlying group often determine the characteristics and error-correcting
capabilities of these codes. The study of perfect codes (subsets of a graph with specific distance properties relevant to error correction)
sometimes involves Cayley graphs of groups.

In such contexts, verifying the group structure ensures that the Cayley graph possesses
the intended properties necessary for code construction and analysis. Additionally, research explores the construction of generator
and parity check matrices for error-correcting codes directly from the Cayley tables of certain algebraic structures \cite{zurekCodingTheory}.

\subsection{Prior Work}
\begin{itemize}
    \item Brute-force: \( O(n^3) \) associativity checks.
    \item Light's \( O(n^2 \log n) \) deterministic method.
    \item Rajagopalan \& Schulman's randomized \( O(n^2 \log(1/\delta)) \).
\end{itemize}

\subsection{Contributions}
\begin{itemize}
    \item First deterministic \( O(n^2) \) algorithm.
    \item Use of \textbf{basis sets} and \textbf{4-associativity}.
    \item Reduction in the search for large subgroups by group decomposition.
\end{itemize}

\section{Technical Overview}
\subsection{Testing Group Axioms Efficiently}
\begin{itemize}
    \item Identity: Check \( \exists e \in G \) s.t. \( e \cdot g = g \cdot e = g \) (\( O(n) \)).
    \item Inverses: Find \( g^{-1} \) via exponentiation (\( O(n \log n) \)).
\end{itemize}

\subsection{4-Associativity and Basis Sets}
\subsubsection{Basis Definition}
A set \( S \subseteq G \) where \( S \cdot S = G \) and \( |S| = O(\sqrt{n}) \).

\subsubsection{Key Lemma}
If \( S \) satisfies 4-associativity, then \( G \) is associative. \\
\textbf{Proof Sketch}: Expand arbitrary triples into basis elements using consistency of quadruples.

\subsection{Reduction to Finding Large Subgroups}
\label{sec:large-subgroups}

%\subsubsection{Overview}
The core challenge in this subsection is constructing a \emph{basis} $S \subseteq G$ with $|S| = O(\sqrt{n})$ and $S^2 = G$, which enables efficient associativity testing via 4-associativity. The authors show that this reduces to finding \emph{large subgroups} $H \leq G$ where $|H| \geq \sqrt{|G|}$.

\begin{definition}[Large Subgroup]
A proper subgroup $H \leq G$ is \textbf{large} if $\sqrt{|G|} \leq |H| < |G|$.
\end{definition}

\begin{definition}[Group Decomposition]
For a group $G$ of size $n$ and parameter $\ell$, an \textbf{$\ell$-decomposition} is a pair $(A,B)$ where:
\begin{itemize}
    \item $A \cdot B = G$
    \item $|A| \leq 2\ell$
    \item $|B| \leq n/\ell$
\end{itemize}
\end{definition}

The reduction proceeds via three main phases:

\textbf{Phase 1: Recursive Decomposition}
Given a large subgroup $H$, we recursively decompose $G$:
\begin{itemize}
    \item \textbf{Base Case}: If $G$ is cyclic of prime order, use arithmetic progressions.
    \item \textbf{Recursive Case}:
    \begin{enumerate}
        \item If $|H| \geq n/(2\ell)$, set $A$ as a left transversal of $H$, $B = H$.
        \item Else, recursively decompose $H$ and lift to $G$ via transversals.
    \end{enumerate}
\end{itemize}

\begin{algorithm}[H]
\caption{GroupDecomposition}
\begin{algorithmic}[1]
%\Require Group $G$, parameter $\ell$
\IF{$|G|$ is prime}
    \STATE Return cyclic decomposition
\ELSE
    \STATE $H \gets \text{LargeSubgroup}(G)$
    \IF{$|H| \geq n/(2\ell)$}
        \STATE $A \gets \text{LeftTransversal}(G,H)$
        \STATE $B \gets H$
    \ELSE
        \STATE $(A',B') \gets \text{GroupDecomposition}(H,\ell)$
        \STATE $A \gets A'$
        \STATE $B \gets B' \cdot \text{RightTransversal}(G,H)$
    \ENDIF
\ENDIF
\RETURN $(A,B)$
\end{algorithmic}
\end{algorithm}

\textbf{Phase 2: Transversal Computation}
\begin{lemma}[Left Transversal]
A left transversal $T$ of $H \leq G$ can be found in $O(n)$ time by:
\begin{enumerate}
    \item Greedily selecting representatives from distinct cosets
    \item Removing entire cosets after selection
\end{enumerate}
\end{lemma}

\textbf{Phase 3: Basis Construction}
\begin{corollary}
Setting $\ell = \sqrt{n/2}$ yields a basis $S = A \cup B$ with $|S| \leq 2\sqrt{2n} < 3\sqrt{n}$.
\end{corollary}


\subsection{Reduction to Simple Groups}
\label{sec:simple-groups}

The algorithm reduces the large subgroup search problem to the case of \emph{simple groups} by:
\begin{itemize}
    \item Finding normal subgroups efficiently
    \item Recursively processing quotient groups
    \item Handling simple groups as base cases
\end{itemize}

\begin{definition}[Simple Group]
A group \( G \) is \textbf{simple} if its only normal subgroups are \( \{e\} \) and \( G \) itself.
\end{definition}

\subsubsection{Identifying normal subgroups}
The algorithm identifies normal subgroups through:

\textbf{Step 1: Small Generating Set}
\begin{itemize}
    \item Compute a generating set \( S \) with \( |S| \leq \log|G| \) in \( \widetilde{O}(n) \) time
    \item Uses greedy expansion of non-generator elements
\end{itemize}

\begin{algorithm}[H]
\caption{Generators}
\begin{algorithmic}[1]
\STATE \( H \gets \{e\} \)
\STATE \( S \gets \emptyset \)
\FOR{each \( a \in A \)}
    \IF{\( a \notin H \)}
        \STATE \( S \gets S \cup \{a\} \)
        \STATE \( H \gets \langle S \rangle \)
    \ENDIF
\ENDFOR
\RETURN \( S \)
\end{algorithmic}
\end{algorithm}

\textbf{Step 2: Conjugacy Class Graph}
\begin{itemize}
    \item Build graph where vertices are conjugacy classes
    \item Edge \( C_i \to C_j \) if elements from \( C_i \) multiply into \( C_j \)
    \item Compute in \( \widetilde{O}(n) \) time using generators
\end{itemize}

\textbf{Step 3: Closed Union Detection}
\begin{itemize}
    \item Normal subgroups correspond to closed unions of classes
    \item Dynamically maintain reachability in the graph
\end{itemize}

\subsubsection{Recursive Reduction}
\begin{lemma}[6.2]
Given an algorithm for simple groups, LargeSubgroup runs in \( O(n^{3/2+\epsilon}) \) time for arbitrary groups.
\end{lemma}

\begin{algorithm}[H]
\caption{LargeSubgroup}
\begin{algorithmic}[1]
\STATE \( N \gets \text{NormalSubgroup}(G) \)
\IF{\( N = G \)}
    \RETURN \(\text{LargeSubgroupOfSimpleGroup}(G)\)
\ELSIF{\( |N| \geq \sqrt{n} \)}
    \RETURN \( N \)
\ELSIF{\( G/N \) has prime order}
    \STATE Find element \( g \) of prime order \( p \)
    \RETURN \( \langle g \rangle \)
\ELSE
    \STATE \( M \gets \text{LargeSubgroup}(G/N) \)
    \RETURN \( M \cdot N \)
\ENDIF
\end{algorithmic}
\end{algorithm}

\subsubsection{Complexity Analysis}
\begin{itemize}
    \item Normal subgroup finding: \( \widetilde{O}(n) \) via dynamic graph maintenance
    \item Recursion depth: \( O(\log n) \) (each step reduces problem size)
    \item Bottleneck: Simple group handling (\( O(n^{3/2+\epsilon}) \))
\end{itemize}

\begin{theorem}
The reduction to simple groups preserves the overall \( O(n^{3/2+\epsilon}) \) time complexity when combined with:
\begin{itemize}
    \item \( \widetilde{O}(n) \) normal subgroup detection
    \item \( O(n^{3/2+\epsilon}) \) simple group processing
\end{itemize}
\end{theorem}

\subsection{Finding Large Subgroups of Simple Groups}
\label{sec:simple-subgroups}

The algorithm handles simple groups through a classification-based approach:

\begin{definition}[Simple Group Types]
Finite simple groups fall into three categories:
\begin{itemize}
    \item Cyclic groups of prime order
    \item Alternating groups $A_n$ ($n \geq 5$)
    \item Groups of Lie type and sporadic groups
\end{itemize}
\end{definition}

\subsubsection{Classification Strategy}
\begin{itemize}
    \item \textbf{Step 1: Group Identification} - Enumerate possible isomorphisms:
    \begin{algorithm}[H]
    \caption{EnumerateGroups}
    \begin{algorithmic}[1]
    \FOR{each simple group family $f$}
        \STATE Solve $|f(m,q)| = n$ for parameters $(m,q)$
        \IF{solution exists}
            \STATE Add $(f,m,q)$ to candidate list
        \ENDIF
    \ENDFOR
    \RETURN candidates
    \end{algorithmic}
    \end{algorithm}
    
    \item \textbf{Step 2: Type-Specific Handling} - Different approaches per category
\end{itemize}

\subsubsection{Algorithmic Approaches}

\textbf{Case 1: Alternating Groups ($A_n$)}
\begin{itemize}
    \item Find the natural subgroup $A_{n-1}$
    \item Implementation steps:
    \begin{enumerate}
        \item Identify all 3-cycles (elements of order 3)
        \item Find a set conjugate to $\{(1,2,k) | 3 \leq k \leq n\}$
        \item Remove one generator to get $A_{n-1}$
    \end{enumerate}
    \item Runtime: $\widetilde{O}(n)$
\end{itemize}

\textbf{Case 2: Groups of Lie Type}
\begin{itemize}
    \item Key tool: Borel subgroups
    \begin{definition}[Borel Subgroup]
    A maximal connected solvable subgroup.
    \end{definition}
    
    \item Implementation:
    \begin{algorithm}[H]
    \caption{BorelSubgroup}
    \begin{algorithmic}[1]
    \STATE Find Sylow $p$-subgroup $P$ (char. $p$ of field)
    \STATE Compute normalizer $B = N_G(P)$
    \RETURN $B$
    \end{algorithmic}
    \end{algorithm}
    
    \item Construct parabolic subgroup:
    \begin{itemize}
        \item Find element $a$ such that $P = \langle B, a \rangle$ is large
        \item Verify $|P| \geq \sqrt{|G|}$
    \end{itemize}
    \item Runtime: $O(n^{3/2+\epsilon})$
\end{itemize}

\textbf{Case 3: Sporadic Groups}
\begin{itemize}
    \item Finite list of 26 exceptions + Tits group
    \item Precomputed large subgroups from group theory literature
    \item Constant-time lookup
\end{itemize}

\subsubsection{Theoretical Guarantees}
\begin{theorem}[Existence]
Every non-prime simple group contains a large subgroup.
\end{theorem}

\begin{proof}[Proof Sketch]
\begin{itemize}
    \item Alternating: $|A_{n-1}| \geq \sqrt{|A_n|}$
    \item Lie type: Parabolic subgroups satisfy size requirement
    \item Sporadic: Verified case-by-case
\end{itemize}
\end{proof}

\subsubsection{Complexity Analysis}
\begin{itemize}
    \item Alternating groups: $\widetilde{O}(n)$ (dominated by 3-cycle detection)
    \item Lie type: $O(n^{3/2+\epsilon})$ (Borel subgroup construction)
    \item Sporadic: $O(1)$ (table lookup)
    \item Classification: $O(\log^4 n)$ (parameter solving)
\end{itemize}

\begin{lemma}
The simple group handling preserves the overall $O(n^{3/2+\epsilon})$ time complexity.
\end{lemma}

% 3. Critical Analysis (3–4 pages)
\section{Critical Analysis}
\subsection{Strengths}
\begin{itemize}
    \item Optimal complexity (\( \Omega(n^2) \) lower bound).
    \item Novel techniques: 4-associativity, group decomposition.
\end{itemize}

\subsection{Limitations}

\subsection{Open Problems}

% 4. Formalni dokazi (1–2 strani)
\section{Formal Proofs}
\subsection{4-Associativity Lemma}

\subsection{Existence of Large Subgroups}
Proof that non-prime finite groups have subgroups of size \( \geq \sqrt{n} \).

% 5. Conclusion (0.5–1 page)
\section{Conclusion}
Summary of results, implications for computational group theory, and future directions.

% References
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
